From physical principles to ???

### 2016-03-30 2015CompleteDerivation/completeDerivation.tex
State variable evolution is now ready for review

Repeatability is also fundamental. First, to make sure the process is indeed correlated to the states. Second, because "a single take does not a measurement make". One has to gather enough statistics. Note that the number of takes influences the outcomes: with greater statistics the precision and number of distinguishable cases increases. Therefore the processes, as we define them, may require repeated interactions with similarly prepared states. They may even be a combination of different kinds of interactions that, taken all together, create a set of distinguishable outcomes. But in the end, however complicated it is, the conceptual model remains the same: each process has a set of possible outcomes, and each possible outcome will be associated with a set of states consistent with that outcome. For example, if the cannonball deformed the ground by this amount then its kinetic energy at impact was within this range; if the electron follows a certain path, then its state is among the ones that has spin $+1/2$.
However precise our measurements are, we can only gather a finite amount of statistics and each outcome is expressed by a finite set of digits; therefore, the set of outcomes is countable.\footnote{The information provided by the process as measured by Shannon's entropy is finite.} Note that different outcomes for the same process can overlap. For example, $4.12 \pm 0.05$ cm and $4.13 \pm 0.05$ cm are both legitimate possible outputs of the same measurement device. But since all states must be distinguishable, given two arbitrary states there must be a process precise enough to tell them apart. That is, the potential outcomes associated with the two states do not overlap. For example, $4.12 \pm 0.0005$ cm and $4.13 \pm 0.0005$ cm do not overlap anymore.

@@

TODO: maybe best to define state vector directly, and state variables as part of state vector? Define discrete state variable when co-domain is integers and define continuous state variable when co-domain is real? Can we prove a state vector always exists, of real variables? For discrete topology, we should always be able to do it. In general, state vector defines charts, so we should not have a single chart for the whole state space. What does it mean physically? Does it matter?

TODO: Still not clear what's the best way to introduce things. We do want to show that state variable are maps from states to "labels". We want to call discrete variables those that map to integer and continuous variables those that map to real numbers. Then we want to concentrate on state space that can be described by a finite number of either discrete variables or continuous variables. State vector may not be valid in general: e.g. we can't assume we have a single chart for the whole space.

\begin{prop}\label{state_vector}
TODO: State vector always exists
\end{prop}

To study time evolution, we need to describe how state variables change under continuous and reversible evolution. There are two ways to do it, and we'll need both. The first approach is to \emph{evolve} the state variables from the initial value $q^i(\mathcal{s})$ to the final value $q^i(\hat{\mathcal{s}})$. The result is a trajectory $q^i(t)$ which is especially useful when state variables correspond to physically meaningful quantities. For example, we track how the temperature or the pressure of an ideal gas changes. Evolved state variables, however, make it hard to find relationships that are invariant and common to all deterministic and reversible processes.

The second approach is to transport the state variables. The idea is to keep the connection to the initial state by labeling the future state, instead by the future values of $q^i$, by the original one. So we introduce a new set of variables for which $\hat{q}^i(\hat{\mathcal{s}})=q^i(\mathcal{s})$, which we can always do as the evolution is deterministic and reversible. The level sets of $\hat{q}^i$ are the evolved level sets of $q^i$, which allows us to study how groups of states evolve in time. Moreover, once we identify an invariant for $q^i$ this is automatically satisfied by $\hat{q}^i$. Given that the value of the transported state variables does not change during evolution, and that it is unique for each initial state, transported state variable provide a way to label the trajectories themselves. While the numeric value of the transported state variables remains the same, the state variable (and its physical meaning) changes. For example, pressure and temperature will be combined depending on the particular process.

Evolved state variables are useful to write equations of motions, study how physical quantities change, form a physical picture of what happens. Transported state variables are useful to write invariants, study state space trajectories, from a geometric picture for the state space. The two, though, are tightly linked. In fact:
\begin{align*}
\Delta q^i &= q^i(\hat{\mathcal{s}}) - q^i(\mathcal{s}) \\
\hat{q}^i(\hat{\mathcal{s}}) &= q^i(\hat{\mathcal{s}}) - \Delta q^i
\end{align*}
Intuitively, the transported variables change in the oppose way as the evolution, so that the value of the variables remain the same.\footnote{One should not confuse evolved/transported state variable with active/passive transformations or with Schroedinger/Heisenberg pictures. In those cases, the choice is whether to change the state or the coordinates/observables. In our case, the state is always changing. The choice is whether to change the state variables or their value.}

\begin{defn}\label{evolved_transported_variable}
TODO: Define evolved and transported variables in terms of pushback/pushforward
\end{defn}

\begin{justification}
TODO: prove existence
\end{justification}

For discrete state variables, the total number of states is the product of the possibilities of each independent state variable. That is: if $\Delta q^1$ and $\Delta q^2$ are range of possibilities for two independent variables, the total number of possibilities $\#(\Delta q^1) \#(\Delta q^2)$. These relationships are maintained during evolution by the transported variables. Since the value remains unchanged, the range of possibilities for each value remain the same: $\#(\Delta \hat{q}^i) = \#(\Delta q^i)$. The variables remain independent as well, all possible combination remain, and we have: $\#(\Delta \hat{q}^1) \#(\Delta \hat{q}^2) = \#(\Delta q^1) \#(\Delta q^2)$. These relationship will later be extended in the continuous case to independent degrees of freedom for a Hamiltonian system, and will justify the symplectic form. Though this can't be done in general and requires further assumptions.

In the case of continuous state, if each state can be identified by n independent state variables then, at least locally, we can chart the state space on $\mathcal{R}$: the state space is a manifold. Studying in general how ranges of possibilities evolve, though, becomes problematic. Time evolution will map points, while range of continuous possibilities will be length. We would need a metric, invariant during the evolution, telling us how ranges of possibilities map from one variable to another, possibly describing different physical quantities. We will not study this case, but we do note that, if such a metric exist, it depends on the system and on the choice of state variables.

\begin{prop}\label{discrete_state_space}
The space state $\mathcal{S}$ for a system fully identified by a set of $n$ independent discrete state variables $q^i : \mathcal{S} \rightarrow \mathbbm{I}^i$ is isomorphic to $\prod\limits_{i=1}^n\mathbbm{I}^i$. Let $I^i \subseteq \mathbbm{I}^i$ be finite sets of possibilities for each variable. $\#(\bigcap\limits_{i=1}^{n}{q^i}^{-1}(I^i))=\prod\limits_{i=1}^{n}\#(I^i)$.
\end{prop}

\begin{proof}
TODO
\end{proof}

TODO: in continuous, not necessarily the Cartesian product. But we can always create a chart with state variables. State space is a manifold of dimension n. In principle, we could say more. For example, a deterministic/reversible map would not map a set to its subset or superset. We could define a metric, this would give us an ergotic system. Not going to do that.

We have seen that physical systems undergoing deterministic and reversible evolution are properly described by a dynamical system. Systems described by a discrete set of states have nice properties for counting possibilities of independent variables, which are maintained during the evolution. System described by a continuous set of states, instead, do not posses in general such properties.

@@

Given that scientific reductionism (i.e. the idea of reducing physical systems and interactions to the sum of their constituent parts in order to make them easier to study) is at the heart of fundamental physics, we should explore how to characterize a system in terms of its components. In general, this is quite a complicated thing to do, that requires intimate knowledge of the system at hand. So we simplify our problem and study an idealized case: one where the components are homogeneous and infinitesimal. What we'll find is that under the additional assumption that the system is infinitesimally reducible (i.e. its state is equivalent to the state of its parts) and that each part undergoes deterministic and reversible evolution, the motion is suitably described by the standard framework of classical Hamiltonian particle mechanics.

>>>>

Repeatability is also fundamental. First, to make sure the process is indeed correlated to the states. Second, because "a single take does not a measurement make". One has to gather enough statistics. Note that the number of takes influences the outcomes: with greater statistics the precision and number of distinguishable cases increases. Therefore the processes, as we define them, may require repeated interactions with similarly prepared states. They may even be a combination of different kinds of interactions that, taken all together, create a set of distinguishable outcomes. But in the end, however complicated it is, the conceptual model remains the same: each process has a set of possible outcomes, and each possible outcome will be associated with a set of states consistent with that outcome. For example, if the cannonball deformed the ground by this amount then its kinetic energy at impact was within this range; if the electron follows a certain path, then its state is among the ones that has spin $+1/2$.

However precise our measurements are, we can only gather a finite amount of statistics and each outcome is expressed by a finite set of digits; therefore, the set of outcomes is countable.\footnote{The information provided by the process as measured by Shannon's entropy is finite.} Note that different outcomes for the same process can overlap. For example, $4.12 \pm 0.05$ cm and $4.13 \pm 0.05$ cm are both legitimate possible outputs of the same measurement device. But since all states must be distinguishable, given two arbitrary states there must be a process precise enough to tell them apart. That is, the potential outcomes associated with the two states do not overlap. For example, $4.12 \pm 0.0005$ cm and $4.13 \pm 0.0005$ cm do not overlap anymore.

@@

To study time evolution, we need to describe how state variables change under deterministic and reversible evolution. There are two ways to do it, and we'll need both. The first approach is to \emph{evolve} the state variables from the initial value $q^i(\mathcal{s})$ to the final value $q^i(\hat{\mathcal{s}})$. The result is a trajectory $q^i(t)=q^i(\lambda(t))$ which is especially useful when state variables correspond to physically meaningful quantities. For example, we track how the temperature or the pressure of an ideal gas changes. Evolved state variables, however, make it hard to find relationships that are invariant and common to all deterministic and reversible processes.

The second approach is to transport the state variables. The idea is to keep the connection to the initial state by labeling the future state, instead by the future values of $q^i$, by the original one. For example, the evolved state variable will not tell us the current value of pressure, but the one for the initial conditions. So we introduce a new set of variables for which $\hat{q}^i(\hat{\mathcal{s}})=q^i(\mathcal{s})$, which we can always do as the evolution is deterministic and reversible. The level sets of $\hat{q}^i$ are the evolved level sets of $q^i$ (e.g. all the states that started with a particular value for pressure) which allows us to study how groups of states evolve in time. Given that the value of the transported state variables does not change during evolution, and that it is unique for each initial state, transported state variables also provide a way to label the trajectories themselves.

Evolved state variables are useful to write equations of motions, study how physical quantities change, form a physical picture of what happens. Transported state variables are useful to write invariants, study state space trajectories, form a geometric picture for the state space.\footnote{One should not confuse evolved/transported state variable with active/passive transformations or with Schroedinger/Heisenberg pictures. In those cases, the choice is between changing the state or the coordinates/observables. In our case, the state is always changing. The choice is between tracking the change using different state variables or different values.}

\begin{defn}\label{evolved_transported_variable}
Let $q^i$ be a set of state variables and $\mathcal{T}_{\Delta t}$ a deterministic and reversible evolution map. The evolved state variables are given by $q^i \circ \mathcal{T}_{\Delta t}$. The transported state variables are given by $q^i \circ \mathcal{T}_{-\Delta t}$.
\end{defn}

To show the usefulness of transported state variables, we can prove the following:

\begin{prop}\label{discrete_state_metric}
	Let $U \subseteq \mathcal{S}$ a set of states fully identified by a set of $n$ independent discrete state variables $q^i$. Let $\Delta q^i = (q^i)^{-1}(U)$ the range of possibilities of each variables. Then $\#(\Delta q^i)=\#(\Delta \hat{q}^i) \; \forall i$ and $\#(U)=\prod\limits_{i=1}^{n}\#(\Delta q^i)=\prod\limits_{i=1}^{n}\#(\Delta \hat{q}^i)=\#(\hat{U})$ where $\hat{U}=\mathcal{T}_{\Delta t}(U)$.
\end{prop}

\begin{proof}
	We claim $\#(\Delta q^i)=\#(\Delta \hat{q}^i)$. For each $\mathcal{s} \in U$, let $\hat{\mathcal{s}}=\mathcal{T}_{\Delta t}(\mathcal{s})$. We have $\hat{q}^i(\hat{\mathcal{s}}) = q^i(\mathcal{s})$. Therefore $\Delta q^i = (q^i)^{-1}(U) = (\hat{q}^i)^{-1}(\hat{U})=\Delta \hat{q}^i$. $\#(\Delta q^i)=\#(\Delta \hat{q}^i)$.

	We claim $\#(U)=\prod\limits_{i=1}^{n}\#(\Delta q^i)$. As $q^i$ are independent variables, the states are the Cartesian product of the possibilities of each variable.
\end{proof}

What happens is that, because the variables are independent, the total number of states is the product of the number of possibilities for each variables. If $\Delta q^1$ and $\Delta q^2$ are ranges of possibilities for two independent variables, the total number of possibilities $\#(\Delta q^1) \#(\Delta q^2)$. By construction, the set of possibilities for each independent variables $\Delta q^i$ is the same as the set of possibilities for the transported variable $\Delta \hat{q}^i$. Also, the transported variables remain independent. Therefore the relationship $\#(\Delta \hat{q}^1) \#(\Delta \hat{q}^2) = \#(\Delta q^1) \#(\Delta q^2)$ is valid throughout the evolution.

We'll see that very similar relationships are what define Hamiltonian and Lagrangian mechanics. But recovering them in the continuous case is trickier, and in fact will be part of the challenge in the following sections. Consider the map $q'=aq$ with $0<a<1$. At first glance, it's a bijective continuous map so we may think it can represent a deterministic and reversible evolution. Yet, $\Delta q = [-b, b] \supset \Delta q'$: a set of states is mapped to a proper subset (i.e. to fewer states) which does not make sense for a reversible process. In the limit where we apply the map an infinite amount of times, any value of $q$ will be brought infinitely close to $0$, which also does not sound like a reversible process. The issue is that while the evolution map will map points to point, not all point to point map can be considered deterministic and reversible. Discrete sets already come with a way to measure and compare the number of elements which is preserved by any bijective map. With continuous sets, we need to do extra work to properly define a measure (or metric) that allows to count states and possibilities.

@@

Given that scientific reductionism (i.e. the idea of reducing physical systems and interactions to the sum of their constituent parts in order to make them easier to study) is at the heart of fundamental physics, we should explore how to characterize a system in terms of its components. That is, we want to study the relationship between the state space $\mathcal{C}$ of the composite system and the state space $\mathcal{S}$ of its parts. In general, this is quite a complicated thing to do, that requires intimate knowledge of the system at hand. So we simplify our problem and study an idealized case: one where the components are homogeneous and infinitesimal. What we'll find is that under the additional assumption that the system is infinitesimally reducible (i.e. its state is equivalent to the state of its parts) and that each part undergoes deterministic and reversible evolution, the motion is suitably described by the standard framework of classical Hamiltonian particle mechanics.
